Recent advances in random matrix theory applied to heavy-tailed multivariate time series

### 2015-09-07 papers/Number1/paper.tex

\subsection{Overview of light-tailed case}\label{sec:light_tail_review}
Yin, Bai and Krishnaiah proved in their 1988 paper
\cite{yin:bai:Krishnaiah:1988} the following result for a $p \times n$
matrix $\bf X$ with iid entries $X_{ij}$ under the assumption $\E
X_{11} = 0$, $\E X_{11}^4 < \infty$, and $q = \lim_{n \to \infty} p/n
> 0$
\[
\lambda_{\M}\left({1 \over n}\bf{XX'}\right) \overset{\as}{\to} \left(
  1 + \sqrt{q}
\right)^2 \E X_{11}^2 \text{ as }n \to \infty
\]
When $\E X_{11} = \infty$, Yin and Silverstein proved in 1988
\cite{yin:silverstein:1988}
\[
\lim_{n \to \infty} \sup \lambda_{\M}\left({1 \over
    n}\bf{XX'}\right) = \infty
\]
Later on, several authors \cite{Pillai:Yin:2011, tao:vu:2012,
  Wang:2012} looked into empirical covariance matrix $S_{n, p} = {1
  \over n} \bf{XX}'$, where the entries of $\bf X$, denoted $X_{ij}$
are independent and satisfy the following condition which is referred
to as condition {\bf C0}.
\begin{enumerate}
\item $\E X_{ij} = 0$, $\E X_{ij}^2 = 1$ and
\item $\forall u > 1, \exists \theta > 0$ such that
  \[
  \P(X_{ij} > u) = \theta^{-1} e^{-u^{\theta}}
  \]
\end{enumerate}
Under condition {\bf C0}, Pillai and Yin \cite{Pillai:Yin:2011} proved
that, for $q = \lim_{n \to \infty} {p \over n} \in (0, \infty)
\setminus \{1\}$, $\forall \zeta > 0, \exists C > 0, \exists \delta >
0$,
\[
\P\left[
  \left|
    \lambda_{(j)}(S_{n,p}) - \gamma_{j}^{n, p}
  \right| > \varphi^{C} p^{-2/3} \tilde{j}^{-1/3}
\right] \leq p^{\delta} \exp(-\varphi^\zeta)
\]
where $\lambda_{(j)}(S_{n,p})$ is the $j$-th largest eigenvalue of
$S_{n,p}$ and
\begin{eqnarray*}
  \varphi &=& (\log p)^{\log \log p} \\
  \tilde j &=& \min\{(p \wedge n) + 1 - j, j\} \\
  \gamma_{j}^{n, p} &=& \text{inf}\left\{\gamma: \int_\gamma^{\lambda_\M}
  \mu_{n,p}(x) dx \leq {j \over p}\right\} \\
  \lambda_\M &=& (1 + \sqrt q)^2 \\
  \lambda_\m &=& (1 - \sqrt q)^2 \\
  \mu_{n,p}(x) &=& {1 \over 2\pi q x} \sqrt{(x - \lambda_\m)(\lambda_\M - x)}
\end{eqnarray*}

Tao and Vu \cite{tao:vu:2012} relaxed the exponential decay condition
of {\bf C0} and replaced it with a high-moment condition, and proved the
four-moment theorem, which was later extended by Wang
\cite{Wang:2012}. In their setting, a $p \times n$ matrix $\bf X$ is
said to obey condition {\bf C1} with constant $C_0$ if $\forall 1
\leq i \leq p, \forall 1 \leq j \leq n$, $\E X_{ij} = 0$, $\E
X_{ij}^2 = 1 $ and $\sup_{i,j} \E|X_{ij}|^{C_0} < C$ for some constant
$C$ independent of $p$ and $n$.

Assume matrices $\bf X$ and $\bf Y$ both satisfy condition {\bf C1}
with constant $C_0$ and in addition they match up to order $4$,
i.e. for all $m, l \geq 0$ such that $m + l \leq 4$, $\E[\mathscr{
R}(X_{ij})^m \mathscr{I}(X_{ij})^l] = \E[\mathscr{R}(Y_{ij})^m
\mathscr{I}(Y_{ij})^l]$. Assume also $\lim_{n \to \infty} p/n = q \in
(0, 1]$. Let $G: \mathbb R^k \to \mathbb R$ be a smooth function
obeying the derivative bounds
\[
\left| \nabla^j G(x) \right| < n^{c_0}
\]
for some $c_0 > 0$. Then the following holds for all $1 \leq i_1 \leq
i_2 \leq \cdots \leq i_{k} \leq p$, $1 \leq k \leq p$:
\[
\left|
\E G(n\lambda_{i_1}, \dots, n\lambda_{i_k}) - 
\E G(n\omega_{i_1}, \dots, n\omega_{i_k})
\right| < n^{- c_0}
\]
where $\lambda_1 \leq \lambda_2, \leq \cdots \leq \lambda_p$ are the
eigenvalues of ${1 \over n} XX'$ and $\omega_1 \leq \omega_2, \leq
\cdots \leq \omega_p$ are those of ${1 \over n} YY'$.

\subsection{Finite $p$: Anderson}
\subsection{Growing $p$: Johnstone, Tao + Vu}

@@

\subsection{$\alpha \in (0,2)$: Overview of existing results: Davis et
  al., Soshnikov, Auffiner et al.}
\subsection{$\alpha \in (0,2)$: Our extension of these results; only
  order statistics of $Z_{it}$'s necessary and no sums (proof by the
  technical result)}
To illustrate the validity of the results, we pick the following
distribution function for $Z$ and compare the eigenvalues with
the order statistics of $D_i = \sum_{t=1}^n Z_{it}^2$ and $Z^2$ using
simulated data.
\[
f(x) = {d \over dx}\P(Z \leq x) = \left\{
  \begin{array}{ll}
    {\alpha \over (4x)^{\alpha + 1}} & |x| > 1/4 \\
    \alpha & \text{ otherwise }
  \end{array}
\right.
\]
This function is shown in figure \ref{fig:densitySim}.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.40]{densitySim.pdf}
  \caption{Density function of simulated data. $\alpha=1.6$ is chosen
    for simulation.}
  \label{fig:densitySim}
\end{figure}
Using the simulated data, empirical density functions are obtained for
$a_{np}^{-2}\lambda_{(i)} - a_{np}^{-2}D_{(i)}$ and
$a_{np}^{-2}\lambda_{(i)} - a_{np}^{-2}Z^2_{(i)}$. These are shown in
figure \ref{fig:lambda_comparison}.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.40]{lambda_comparison.pdf}
  \caption{A total of 20000 pairs are compared in each case.}
  \label{fig:lambda_comparison}
\end{figure}

>>>>

\subsection{Overview of light-tailed
  case}\label{sec:light_tail_review}
As early as in 1962, T. W. Anderson \cite{anderson:1963} showed that,
when the entries of the data matrix has normal distribution ($\E
X_{ij} = 0$ for $i,j = 1,2,\dots, p$; $\E X_{ii}^2 = 2\sigma^2, \E
X_{ij}^2 = \sigma^2$ for $i \neq j$), the joint density function
of the eigenvalues of the sample covariance matrix is given by
\[
f(\lambda_1, \lambda_2, \dots, \lambda_p; \sigma, p) \propto \prod_{i=1}^p \exp\left(
  -{\lambda_i^2 \over 4\sigma^2}
\right) \prod_{i < j} (\lambda_i - \lambda_j)
\]
In 1988, Yin, Bai and Krishnaiah proved in
\cite{yin:bai:Krishnaiah:1988} the following result for a $p \times n$
matrix $\bf X$ with iid entries $X_{ij}$ under the assumption $\E
X_{11} = 0$, $\E X_{11}^4 < \infty$, and $q = \lim_{n \to \infty} p/n
> 0$
\[
\lambda_{\M}\left({1 \over n}\bf{XX'}\right) \overset{\as}{\to} \left(
  1 + \sqrt{q}
\right)^2 \E X_{11}^2 \text{ as }n \to \infty
\]
When $\E X_{11}^4 = \infty$, Yin and Silverstein proved in 1988
\cite{yin:silverstein:1988}
\[
\lim_{n \to \infty} \sup \lambda_{\M}\left({1 \over
    n}\bf{XX'}\right) = \infty
\]
Later in 2001, I. M. Johnstone showed that, when the data matrix has
standard normal distribution, i.e. $X_{ij} \sim N(0, 1)$, the largest
eigenvalue of the sample covariance matrix follows the Tracy-Widom
distribution:
\[
{ł_1 - \mu_{np} \over \sigma_{np}} \overset{d}{\to} W_1 \sim F_1
\text{ as }{n, p \to \infty}
\]
where $\mu_{np}$ and $\sigma_{np}$ are constants dependent on $p$ and
$n$ and the distribution function $F_1$ is given by
\[
F_1(s) = \exp\left\{
  -{1 \over 2} \int_{s}^\infty \left[
    q(x) + (x - s) q^2(x)
  \right] dx
\right\}
\]
where $q(x)$ is the unique solution to the Painlev\'e II differential
equation
\begin{eqnarray*}
  q''(x) &=& xq(x) + 2 q^3(x) \\
  q(x) &\sim& \text{Ai}(x) \text{ as } x \to \infty
\end{eqnarray*}
Figure \ref{fig:normal-TW} illustrates the density function of the largest
eigenvalue from a simulated sample and that of the Tracy-Widom.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.5]{normal-TW.pdf}
  \caption{Sample density function of the largest eigenvalue compared
    with the Tracy-Widom density function. The data matrix has
    dimension $200 \times 1000$. An ensamble of 2000 matrices are
    simulated.}
  \label{fig:normal-TW}
\end{figure}

Later on, several authors \cite{Pillai:Yin:2011, tao:vu:2012,
  Wang:2012} looked into sample covariance matrix $S_{n, p} = {1
  \over n} \bf{XX}'$, where the entries of $\bf X$, denoted $X_{ij}$
are independent and satisfy the following condition which is referred
to as condition {\bf C0}.
\begin{enumerate}
\item $\E X_{ij} = 0$, $\E X_{ij}^2 = 1$ and
\item $\forall u > 1, \exists \theta > 0$ such that
  \[
  \P(X_{ij} > u) = \theta^{-1} e^{-u^{\theta}}
  \]
\end{enumerate}
Under condition {\bf C0}, Pillai and Yin \cite{Pillai:Yin:2011} proved
that, for $q = \lim_{n \to \infty} {p \over n} \in (0, \infty)
\setminus \{1\}$, the probability of the $j$-th largest eigenvalue of
the sample covariance matrix deviating significantly from its
theoretical value computed as the $1 - j/p$ quantile of the
Marcenko-Pastur distribution decreases exponentially\footnotemark.

\footnotetext{
In precise terms, this means $\forall \zeta > 0, \exists C > 0,
\exists \delta > 0$,
\[
\P\left[
  \left|
    \lambda_{(j)}(S_{n,p}) - \gamma_{j}^{n, p}
  \right| > \varphi^{C} p^{-2/3} \tilde{j}^{-1/3}
\right] \leq p^{\delta} \exp(-\varphi^\zeta)
\]
where $\lambda_{(j)}(S_{n,p})$ is the $j$-th largest eigenvalue of
$S_{n,p}$ and
\begin{eqnarray*}
  \varphi &=& (\log p)^{\log \log p} \\
  \tilde j &=& \min\{(p \wedge n) + 1 - j, j\} \\
  \gamma_{j}^{n, p} &=& \text{inf}\left\{\gamma: \int_\gamma^{\lambda_\M}
  \mu_{n,p}(x) dx \leq {j \over p}\right\} \\
  \lambda_\M &=& (1 + \sqrt q)^2 \\
  \lambda_\m &=& (1 - \sqrt q)^2 \\
  \mu_{n,p}(x) &=& {1 \over 2\pi q x} \sqrt{(x - \lambda_\m)(\lambda_\M - x)}
\end{eqnarray*}
}

Tao and Vu \cite{tao:vu:2012} relaxed the exponential decay condition
of {\bf C0} and replaced it with a high-moment condition, and proved the
four-moment theorem, which was later extended by Wang
\cite{Wang:2012}. In their setting, a $p \times n$ matrix $\bf X$ is
said to obey condition {\bf C1} with constant $c_0$ if $\forall 1
\leq i \leq p, \forall 1 \leq j \leq n$, $\E X_{ij} = 0$, $\E
X_{ij}^2 = 1 $ and $\sup_{i,j} \E|X_{ij}|^{c_0} < C$ for some constant
$C$ independent of $p$ and $n$.

Assume matrices $\bf X$ and $\bf Y$ both satisfy condition {\bf C1}
with constant $c_0$ and in addition they match up to order $4$,
i.e. for all $m, l \geq 0$ such that $m + l \leq 4$, $\E[\mathscr{
R}(X_{ij})^m \mathscr{I}(X_{ij})^l] = \E[\mathscr{R}(Y_{ij})^m
\mathscr{I}(Y_{ij})^l]$ for all $1 \leq i,j \leq p$. Assume also
$\lim_{n \to \infty} p/n = q \in (0, 1]$. Let $G: \mathbb R^k \to
\mathbb R$ be a smooth function obeying the derivative bounds
\[
\left| \nabla^j G(\vec x) \right| < n^{c_0}
\]
for some $c_0 > 0$ and for all $\vec x \in \mathbb R^k$ and all $0
\leq j \leq 5$. Then the following holds for all $1 \leq i_1 \leq i_2
\leq \cdots \leq i_{k} \leq p$, $1 \leq k \leq p$:
\[
\left|
\E G(\lambda_{i_1}, \dots, \lambda_{i_k}) - 
\E G(\omega_{i_1}, \dots, \omega_{i_k})
\right| < n^{- c_0}
\]
where $\lambda_1 \leq \lambda_2, \leq \cdots \leq \lambda_p$ are the
eigenvalues of $\bf XX'$ and $\omega_1 \leq \omega_2, \leq
\cdots \leq \omega_p$ are those of $\bf YY'$.

Moreover, even when ${\bf XX'}$ and ${\bf YY'}$ only match to order
3, the above inequality still holds provided that the bounds on the
derivatives are strengthened to
\[
|\nabla^j G(\vec x)| < n^{-j c_1}
\]
for all $0 \leq j \leq 5$, all $\vec x \in \mathbb R^k$ and some $c_1$
that is sufficiently large compared to $c_0$.

@@

\subsection{$\alpha \in (0,2)$: Overview of existing results: Davis et
  al., Soshnikov, Auffiner et al.}
\subsection{$\alpha \in (0,2)$: Our extension of these results; only
  order statistics of $Z_{it}$'s necessary and no sums (proof by the
  technical result)}
To illustrate the validity of the results, we pick the following
distribution function for $Z$ and compare the eigenvalues with
the order statistics of $D_i = \sum_{t=1}^n Z_{it}^2$ and $Z^2$ using
simulated data.
\[
f(x) = {d \over dx}\P(Z \leq x) = \left\{
  \begin{array}{ll}
    {\alpha \over (4|x|)^{\alpha + 1}} & |x| > 1/4 \\
    \alpha & \text{ otherwise }
  \end{array}
\right.
\]
This function is shown in figure \ref{fig:densitySim}.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.40]{densitySim.pdf}
  \caption{Density function of simulated data. $\alpha=1.6$ is chosen
    for simulation.}
  \label{fig:densitySim}
\end{figure}
Using the simulated data, empirical density functions are obtained for
$a_{np}^{-2}\lambda_{(i)} - a_{np}^{-2}D_{(i)}$ and
$a_{np}^{-2}\lambda_{(i)} - a_{np}^{-2}Z^2_{(i)}$. These are shown in
figure \ref{fig:lambda_comparison}.
\begin{figure}[htb!]
  \centering
  \includegraphics[scale=0.40]{lambda_comparison.pdf}
  \caption{A total of 20000 pairs are compared in each case.}
  \label{fig:lambda_comparison}
\end{figure}
